The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/extra_numerical_convergence.tex
full draft of appendicies, improved formatting

The two other key numerical parameters in \gadget{} are the tree-force
and time-step accuracy parameters. Specifically, \gadget{} opens a node
if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
the particle is \(r\), side-length is \(l\), and the particle's total
acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
adaptive time stepping with integration accuracy set to \(\eta=0.01\).
(particles must take time-steps smaller than
\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)).

The top panel of Fig.~\ref{fig:methods_convergence} illustrates the
influence of softening length for simulations of Scl's \smallperi{}
orbit with \(10^5\) particles. We show a model with the fiducial
softening length (\(h=0.14\,\kpc\)), and larger and smaller softenings
by a factor \(\sqrt{10}\). The larger softening length is consistent
with Eq.~\ref{eq:power_softening}'s prediction. With a larger softening
length, the halo diverges from the expectation (using \(100\times\) more
particles) across most radii. On the other hand, the fiducial and
smaller softening lengths predict similar final results, although the
smaller softening length simulation deviates slightly more than the
fiducial, likely a result of more collisional encounters. As computation
time increases moderately with decreasing softening length, the fiducial
softening length balances efficiency and accuracy for this simulation.
While smaller than \citet{power+2003}'s suggestion, our choice is
designed for more compact, dwarf satellite halos instead of large,
galaxy-mass isolated halos.

The lower panel of Fig.~\ref{fig:methods_convergence} tests changes to
these parameters. We include a model with smaller integration accuracy
(``small timestep'') and a stricter tree-force tolerance (``high acc.
force''). More precise tolerances on these parameters does not affect
the evolution within the uncertainties of the final profile.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
\caption[Isolation method convergence]{A comparison of the final
profiles using different simulation methods for Sculptor's \smallperi{}
model. The benchmark model is our fiducial \(10^7\) particle run, and
all other models use \(10^5\) paritcles, with their ``converged radius''
marked by the black arrow and softening by the vertical bar.
\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
lengths. \textbf{Bottom:} Models with more precise timestep accuracy
(\(\eta=0.001\) compared to fiducial \(\eta=0.005\)) and gravitational
force accuracy (\(\alpha=0.003\) versus \(\alpha=0.01\)).
\textbf{Summary:} Except for the model with a larger softening length,
all simulations agree within uncertainties. \textbf{Todo: softening
along lower axis always\ldots, changing
linewidth?}}\label{fig:methods_convergence}
\end{figure}

>>>>

The top panel of Fig.~\ref{fig:methods_convergence} illustrates the
influence of softening length for simulations of Scl's \smallperi{}
orbit with \(10^5\) particles. We show a model with the fiducial
softening length (\(h=0.14\,\kpc\)), and larger and smaller softenings
by a factor \(\sqrt{10}\). The larger softening length is consistent
with Eq.~\ref{eq:power_softening}'s prediction. With a larger softening
length, the halo diverges from the expectation (using \(100\times\) more
particles) across most radii. On the other hand, the fiducial and
smaller softening lengths predict similar final results. As computation
time increases with decreasing softening length, the fiducial softening
length balances efficiency and accuracy for this simulation.

The two other key numerical parameters in \gadget{} are the tree-force
and time-step accuracy parameters. Specifically, \gadget{} opens a node
if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
the particle is \(r\), side-length is \(l\), and the particle's total
acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
adaptive time stepping with integration accuracy set to \(\eta=0.01\).
(particles must take time-steps smaller than
\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)). The
lower panel of Fig.~\ref{fig:methods_convergence} tests changes to these
parameters. We include a model with smaller integration accuracy
(\(\eta=0.003\) in ``small timestep'') and a stricter tree-force
tolerance (\(\alpha=0.001\), ``high acc. force''). More precise
tolerances on these parameters does not affect the evolution within the
uncertainties of the final profile.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
\caption[Isolation method convergence]{A comparison of the final
profiles using different simulation methods for Sculptor's \smallperi{}
model. The benchmark model is our fiducial \(10^7\) particle run, and
all other models use \(10^5\) paritcles, with their ``converged radius''
marked by the black arrow and softening by the vertical bar(s).
\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
lengths than the fiducial. \textbf{Bottom:} Models with more precise
timestep accuracy and gravitational force accuracy. \textbf{Summary:}
Except for the model with a larger softening length, all simulations
agree within uncertainties.}\label{fig:methods_convergence}
\end{figure}

### 2025-10-06 thesis/extra_results.tex
full draft of appendicies, improved formatting

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
ellipsoidal models. While each of these variables influences resulting
dark matter evolution, the final stellar component and dark matter
remnant remains largely across alternate initial conditions. We conclude
that consideration of these effects likely would not change our
conclusions.

The following models aim to reproduce the velocity dispersion
\(\lesssim 1\,\kms\) with a similar present-day half-light radius.

@@

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_halo_boundmass.pdf}}
\caption[Tidal dependence on halo concentration]{A comparison of the
evolution of different N-body models for different halo concentrations
(heavier and lighter halo), and the mean orbit instead of the
\smallperi{} orbit. In the left, the maximum circular velocity \(\vmax\)
of the dark matter halo is plotted as a function of simulation time. In
the right, \(\vmax\) is instead plotted as a function of
\(\rmax\).}\label{fig:tidal_tracks_concentration}
\end{figure}

\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}

Figure: The evolution of velocity anisotropy.

@@

Many dwarf galaxies appear to have dark matter cores. In Scl, the
presence or absence of a core has been debated
\citep[e.g.,][]{battaglia+2008, walker+2009a, agnello+evans2012, breddels+helmi2013, amorisco+zavala+deboer2014, richardson+fairbairn2014}.
To simulate the evolution of a cored dark matter halo, we adopt a
``cored-NFW'' model, \begin{equation}{
\rho/\rho   _s = \frac{1}{(1+r/r_s)^2 (r_c/r_s + r/r_s)},
}\end{equation} where \(r_c\) is the core radius. We set
\(M_s=0.79\times10^{10}\,\Mo\) where \(M_s = 4\pi/3\ r_s^3 \rho_s\),
\(r_s=2.76\,\kpc\), \(r_c=0.1r_s\), and use the same truncation as our
fiducial halo (Eq.~\ref{eq:trunc_nfw}). While this is not the largest
possible core, the core is similar in size to the stellar component.
Larger cores would require a more massive halo in order to match the
velocity dispersion at a constant scale radius.

In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
cored model as compared to a heavier halo. The cored halo loses more
total dark matter mass, evolving further along the tidal track as
expected. However, in order to reproduce the observed velocity
dispersion, the final halo reaches a \(\rmax\) and \(\vmax\) as the
fiducial halo. A cored dark matter halo likely cannot be completely
tidally disrupted and still recover the observed velocity dispersion.

\pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}

Figure: Cored versus cuspy initial and final density (todo, fix)..

@@

Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
violating our assumption of spherical symmetry
\citep[e.g.,][]{an+koposov2022}.

We create an oblate initial dark matter halo using \agama{}. The initial
snapshot is sampled from the distribution function \begin{equation}{
\begin{split}
f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \Agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor 0.5 in the \(z\) axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

We find the evolution of the oblate halo to be nearly identical to the
spherical halo, similar to \citet{battaglia+sollima+nipoti2015}. The
oblate halo however becomes spherical as time progresses. If Scl's dark
matter halo is indeed elliptical, this model may not be an adequate
description. Possibly, internal rotation, alternative density structure,
or velocity anisotropy play a role in the stability of oblate initial
conditions to tidal stripping. A full investigation of these effects is
left to future work.

\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_density_i_f.pdf}}

Slices of the oblate density before and after tidal evolution.
\textbf{TODO: isodensity contours on RHS projected major/minor axis}.

@@

In the main text, we focus on recent tidal impacts on Scl and UMi. But,
due to the long-term orbital uncertainties of Scl and UMi with an LMC
(see Fig.~\ref{fig:scl_umi_orbit_uncert}), tidally extreme orbits are
entirely possible. Here, we consider a few examples of extreme orbital
evolution in the long term case. While the actual orbit of either galaxy
almost certainly deviates from these examples (discussion in
Section~\ref{sec:orbital_uncertainties}), these orbits still represent
extreme past tidal histories.

For a more massive LMC (e.g., the L3M11 model), Scl may have undergone
an extreme pericentric passage with the MW \(\sim 6\,\Gyr\) ago. We find
that the 3\(\sigma\) smallest pericentre is \(4\,\kpc\). To test if this
model forms a different density profile, we simulate this model.

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre.
Table~\ref{tbl:orbit_adjustments} records the iterations we take. Our
final model is able to approximately reproduce the observed position of
Scl, but more refinement is required for a closer match.

@@

\section{Present-day properties}\label{present-day-properties}

We find that this model also does not display strong signs of tidal
signatures. Fig.~\ref{fig:scl_mw_impact_i_f} shows the initial and final
stellar profiles for this model (integrated with a lower-resolution
\(10^6\) particles). The model evolves little more than the
\smallperi{}. Given that the Jacobi radius would be within \(R_h\), why
does this model not show strong density perturbations?

We propose that th case Scl undergoes a small MW pericentre is an
impulsive encounter. Because Scl's apocentre is still \(\sim 300\,\kpc\)
in this model, the galaxy passes through the MW with velocity
\(600\,\kms?\). Since the crossing time at the half-light radius is
about 90 Myr, in one crossing-time, the model moves from pericentre to a
Galactocentric distance of \(\sim20\,\kpc\), where tides forced have
dropped substantially. As a result, after \(6\Gyr\) of relatively
isolated evolution, the break radius falls well outside the observed
density profile, but the injection of energy uniformly throughout the
halo only results in moderate DM loss predominantly in the outskirts,
not a reshaping of the inner density profile. Note that the Jacobi
radius is derived on circular orbits, not highly elliptical orbits like
this case.

However, we only note that the stars appear to be consistent but are
well below the \citet{power+2003}'s numerically converged radius.

@@

As we predict that any tidal tails formed in Scl and UMi would be
extremely faint, we deferred discussion of the properties of any such
tails to this section. We note that in extreme cases, \(\sim 2\%\) of
stars may become unbound in either galaxy, forming tidal tails which may
be detectible with substantially better observations.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_sim_stream.pdf}}
\caption[Ursa Minor predicted stream]{The properties of the stream
around the UMi \smallperi{} orbit with Plummer stars. The panels are all
as a function of \(\xi'\), the distance along the stream as defined by
the current GSR proper motion vector. The top panels show the GSR proper
motions in RA and Dec, and the bottom two show the distance and GSR
radial velocities. To sample the stream, we randomly draw 100,000
samples from the snapshot based on the stellar weights. A detectible
gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
the stream is tracked across several
degrees.}\label{fig:umi_tidal_stream}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
\caption[Scl predicted stream]{The predicted properties of a tidal tail
in the Scl model. \textbf{TODO: titles}}\label{fig:scl_tidal_stream}
\end{figure}

>>>>

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
ellipsoidal models. While each of these variables influences total dark
matter evolution, the resulting inner structure is similar across
models. We conclude that consideration of these effects likely would not
change our conclusions.

@@

The following models aim to reproduce the stellar velocity dispersion of
Scl to within \(\lesssim 1\,\kms\) with a similar present-day half-light
radius.

@@

The heavier halo diverges from the initial point orbit more
substantially. We show the results of correcting the orbit (akin to Ursa
Minor's correction, see Section~\ref{sec:orbit_corrections}) in
Fig.~\ref{fig:tidal_tracks_concentration} as the ``heavier, new orbit''
halo. Since correcting the orbit has a relatively small effect on the
tidal evolution, we neglect these corrections in further comparisons for
simplicity.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_halo_boundmass.pdf}}
\caption[Tidal dependence on halo concentration]{A comparison of the
evolution of different N-body models for different halo concentrations
(heavier and lighter halo), and the heavy halo with the action-angle
corrected orbit. In the left, the maximum circular velocity \(\vmax\) of
the dark matter halo is plotted as a function of simulation time. In the
right, \(\vmax\) is instead plotted as a function of
\(\rmax\).}\label{fig:tidal_tracks_concentration}
\end{figure}

@@

Many dwarf galaxies appear to have dark matter cores. In Scl, the
presence or absence of a core has been debated
\citep[e.g.,][]{battaglia+2008, walker+2009a, agnello+evans2012, breddels+helmi2013, amorisco+zavala+deboer2014, richardson+fairbairn2014}.
To simulate the evolution of a cored dark matter halo, we adopt a
``cored-NFW'' model, \begin{equation}{
\rho/\rho   _s = \frac{1}{(1+r/r_s)^2 (r_c/r_s + r/r_s)},
}\end{equation} where \(r_c\) is the core radius. We set
\(M_s=0.54\times10^{10}\,\Mo\) where \(M_s = 4\pi/3\ r_s^3 \rho_s\),
\(r_s=1.08\,\kpc\), \(r_c=r_s\), and use the same truncation as our
fiducial halo (Eq.~\ref{eq:trunc_nfw}). We plot the initial and final
density, as compared to a cuspy NFW, in Fig.~\ref{fig:cored_i_f}.

In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
cored model as compared to a heavier halo. The cored halo loses more
total dark matter mass, evolving further along the tidal track as
expected. However, in order to reproduce the observed velocity
dispersion, the final halo maintains a higher \(\vmax\) than the
fiducial halo.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}
\caption[Cored initial and final density profile]{The initial (dotted
orange) and final (solid orange) 3D density profiles for the cored model
of Scl on the \smallperi{} orbit. The blue thin line represents an NFW
halo with the same \(M_s\) and \(r_s\).}\label{fig:cored_i_f}
\end{figure}

\subsection{Velocity Anisotropy}\label{velocity-anisotropy}

Radial velocity anisotropy may cause halos to disrupt faster
\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
velocity anisotropy, we initialize a model with a velocity anistropy
with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
to \(\beta=1\) and infinity, with scale length \(4r_s\).

Fig.~\ref{fig:anisotropy_i_f} shows the initial and final anisotropy
profiles. The dwarf galaxy becomes more isotropic with tidal evolution.
Particles on more radially-anisotropic orbits are more easily stripped
as they have larger apocentres than more circular orbits of the
equivalent energy. Regardless, the overall tidal evolution is very
similar to our fiducial, isotropic case (see
Fig.~\ref{fig:tidal_tracks_structure}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
profiles for the initially-anisotropic model of Scl on the \smallperi{}
orbit. The final profile is far more isotropic (\(\beta=0\)) than the
initial.}\label{fig:anisotropy_i_f}
\end{figure}

@@

Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
violating the assumption of spherical symmetry
\citep[e.g.,][]{an+koposov2022}. The shape of the underlying dark matter
halo is unknown, however.

We create an oblate initial dark matter halo using \agama{}.
Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after 5Gyr in
isolation) isodensity contours of our model. The initial snapshot is
sampled from the distribution function \begin{equation}{
\begin{split}
f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor 0.5 in the \(z\) axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

We find the evolution of the oblate halo to be nearly identical to the
spherical halo (see Fig.~\ref{fig:tidal_tracks_structure}), similar to
\citet{battaglia+sollima+nipoti2015}. The oblate halo however becomes
spherical after tidal evolution (Fig.~\ref{fig:oblate_i_f}). If Scl's
dark matter halo is indeed elliptical today, this model may not be an
adequate description.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo before and after tidal evolution]{The initial and
final (after 9Gyr of tidal evolution) projected density profiles for the
oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
\(z'\) the major and minor axes. The contours are drawn assuming normal
smoothing of 0.4 kpc and are log-spaced with intervals of 0.105
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

As discussed in Section~\ref{sec:scl_umi_orbit_uncert}, the long-term
orbit of Scl is uncertain. For a more massive LMC (e.g., the L3M11
model), Scl may have undergone an extreme pericentric passage with the
MW \(\sim 6\,\Gyr\) ago. We find that the 3\(\sigma\) smallest
pericentre is \(4\,\kpc\), and simulate this model to test if such a
pericentre may be sufficient.

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, is able to approximately reproduce
the observed position of Scl.

Fig.~\ref{fig:tidal_tracks_umi} compares the tidal evolution of Scl on
the mean, \texttt{MW\ impact} , and \smallperi{} orbit. The mean orbit
evloves less down the tidal track than the \smallperi{} model. Instead,
the \texttt{MW\ impact} orbit experiences most tidal evolution during
its first MW-pericentre. While evolving further along the tidal track,
the stars of this model nevertheless remain exponential
(Fig.~\ref{fig:scl_mw_impact_i_f}). We suggest that the impulsive
pericentric passage does not occur for long enough in this model to
produce the expected extended density profile. A yet more extreme
orbital history would be necessary to tidally transform Scl's stars.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_impact_i_f.pdf}}
\caption[Sculptor MW-impact density profiles]{Similar to
Fig.~\ref{fig:scl_smallperi_i_f} except for the orbit of Scl passing
through the MW.}\label{fig:scl_mw_impact_i_f}
\end{figure}

@@

As discussed in Section~\ref{sec:results}, any hints of a possible tidal
stream around Scl and UMi are beyond the reach of current observational
facilities. However, we can still predict the properties of such a
stream.

Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream}
shows the resulting distributions of velocities and distance along the
stream orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper
motions (mostly for UMi, of \(\sim 0.1\,\masyr\)). However, detecting
such a gradient would require tracing stars across several degrees on
the sky.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
\caption[Sculptor predicted stream]{The predicted properties of a tidal
tail in the Scl model. The panels are all as a function of \(\xi'\), the
distance along the stream as defined by the current GSR proper motion
vector. The top panels show the GSR proper motions in RA and Dec, and
the bottom two show the distance and GSR radial velocities. To sample
the stream, we randomly draw 100,000 samples from the snapshot based on
the stellar weights. A detectible gradient in \(\mu_{\alpha*}\) and LOS
velocity should be detectible if the stream is tracked across several
degrees.}\label{fig:scl_tidal_stream}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_sim_stream.pdf}}
\caption[Ursa Minor predicted stream]{The properties of the stream
around the UMi \smallperi{} orbit with Plummer
stars.}\label{fig:umi_tidal_stream}
\end{figure}

### 2025-10-06 thesis/extra_rv_models.tex
full draft of appendicies, improved formatting

We compile LOS velocities from several spectroscopic surveys. For
Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
\citet{sestito+2023a}; and \apogee{} \citep[DR17,][]{abdurrouf+2022}.
For Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
\citet{sestito+2023b}; and \apogee{}. We then crossmatch all catalogues
to J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we
match to the nearest star within 1--3 arcseconds. We combine
measurements of the same star using inverse-variance weighting.

To reduce likely binaries, we remove stars with significant velocity
dispersions.\footnote{Specifically, using that
  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
  distribution with \(N-1\) measurements.}

We build on J+24's likelihood by adding multiplicative terms in the
total likelihood for the velocity consistency (see
Section~\ref{sec:extra_density}). We assume that the satellite and
background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
halo/background velocity dispersion of a constant
\(\sigma_{\rm halo} = 100\,\kms\) with mean 0
\citep[e.g.][]{brown+2010}. We select stars with velocity-informed
satellite membership probabilities of greater than 0.2. For Scl, we find
1918 unique members and UMi, 831.

For UMi, we shifted the velocities of \citet{spencer+2018}
(\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
a systematic velocity offset. Otherwise all studies appear to agree.

We correct the velocities for the solar motion and on-sky size of the
galaxy. We transform the velocities into the galactic standard of rest
(GSR, i.e.~same location as ICRS but velocities relative to the galactic
centre), and then transform velocities to correct for the apparent
gradient induced by the dwarf's proper motion.\footnote{Specifically,
  sub} We define \(v_{\rm gsr}'\) to be velocities in the The correction
from both effects induces an apparent gradient of about \(1.3\,\kmsdeg\)
for Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. The uncertainty on this velocity
correction are less than the individual star uncertainties and our
derived velocity gradient.

@@

We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
velocity, velocity dispersion, and possible gradient. We assume that the
galaxy follows a planar velocity gradient. The mean velocity at a given
point on the sky, \(\mu\), is assumed to be,

\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
}\end{equation}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[LOS velocity fit to Scl]{Velocity histogram of Scl and UMi in
terms of projected-correction GSR LOS velocity (Eq.~\ref{eq:v_z}). Black
points with error bars are from the crossmatched observed sample, and
green lines represent MCMC fits to the velocity dispersion.}
\end{figure}

Figure: The observed velocity dispersion gradient (black) in 10 equal
number bins and the derived slopes from the model fitting. Scl shows
moderate evidence for an increasing velocity dispersion with radius.
UMi's evidence for a radial gradient is weaker, and Scl's velocity
dispersion appears to be flat outside of \(R_h\).

@@

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
gradient of \(4.3\pm1.3\,\kmsdeg\) at a position angle of
\(-149_{-13}^{+17}\) degrees, similar to past detections
\citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
see also][]{strigari2010}; \citealp{martinez-garcia+2023}}.

@@

We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
dispersion of \(8.8\pm0.2\,\kms\) for UMi. We do not find evidence for a
velocity gradient. This is consistent with previous work
\citetext{\citealp{pace+2020}; \citealp[somewhat
with][]{spencer+2018}; \citealp{martinez-garcia+2023}}.

>>>>

We compile LOS velocities from several spectroscopic surveys. For
Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
\citet{sestito+2023a}; and APOGEE \citep[DR17,][]{abdurrouf+2022}. For
Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
\citet{sestito+2023b}; and APOGEE. We then crossmatch all catalogues to
J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we match
to the nearest star within 1--3 arcseconds. We combine measurements of
the same star using inverse-variance weighting. To reduce likely
binaries, we remove stars with significant velocity
dispersions.\footnote{Specifically, using that
  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
  distribution with \(N-1\) measurements.}

We build on J+24's likelihood by adding multiplicative terms in the
total likelihood for the velocity consistency (see
Section~\ref{sec:the_algorithm}). We assume that the satellite and
background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
halo/background velocity dispersion of a constant
\(\sigma_{\rm halo} = 100\,\kms\) with mean 0
\citep[e.g.][]{brown+2010}. We select stars with velocity-informed
satellite membership probabilities of greater than 0.2. For Scl, we find
1918 unique members and UMi, 831.

For UMi, we shifted the velocities of \citet{spencer+2018}
(\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
a systematic velocity offset. Otherwise, all studies appear to be on a
similar velocity scale.

We correct the velocities for the solar motion and on-sky size of the
galaxy. We transform the velocities into the galactic standard of rest
(GSR, i.e.~same location as ICRS but velocities relative to the galactic
centre), and then transform velocities to correct for the apparent
gradient induced by the dwarf's proper motion \citep[see
also][]{WMO2008, strigari2010}. We define \(v_{\rm gsr}'\) to be
velocities in the GSR frame subtracting the PM-induced gradient. The
correction from both effects induces an apparent gradient of about
\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor. The uncertainty
on this velocity correction are less than the individual star
uncertainties and any derived velocity gradients.

@@

We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
velocity, velocity dispersion, and possible gradients in velocities or
velocity dispersions. We assume that the galaxy follows a planar
velocity gradient. The mean velocity at a given point on the sky,
\(\mu\), is assumed to be,

\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta,
}\end{equation}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[LOS velocity fit to Scl]{\textbf{Left}: Velocity histogram of
Scl and UMi in terms of projected-correction GSR LOS velocity. Black
points with error bars are from the crossmatched observed sample, and
green lines represent MCMC fits to the velocity dispersion.
\textbf{Right} The observed velocity dispersion gradient (black points)
in 10 equal number bins and the derived slopes for MCMC samples (green
lines). Scl shows moderate evidence for an increasing velocity
dispersion with radius, but Scl's velocity dispersion appears to flatten
outside of \(R_h\). UMi's evidence for a radial gradient is
weaker.}\label{fig:rv_hists}
\end{figure}

@@

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\) (see Fig.~\ref{fig:rv_hists}).
Our values are very consistent with previous work
\citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
gradient of \(4.3\pm1.3\,\kmsdeg\) at a position angle of
\(-149_{-13}^{+17}\) degrees, similar to past detections
\citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
see also][]{strigari2010}; \citealp{martinez-garcia+2023}}.

@@

We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
dispersion of \(8.8\pm0.2\,\kms\) for UMi (see Fig.~\ref{fig:rv_hists}).
We do not find evidence for a velocity gradient. This is consistent with
previous work \citetext{\citealp{pace+2020}; \citealp[somewhat
with][]{spencer+2018}; \citealp{martinez-garcia+2023}}.
